---
title: "Setting up the environment"
author: "Bizovi Mihai"
format: html
---

For all:

0. Git + Github account
1. Linux Subsystem activation on Windows
2. R version >= 4.x.x
3. Recent-ish Rstudio >= 2022-01-01
4. [Quarto](https://quarto.org/) - backwards compatible with RMarkdown!
5. `renv` - dependency manager | lock dependencies, without help in own packages
  - create a new RStudio project in the directory you're working
  - project_name.Rproj -> project setting -> use environments

```{r}
renv::init()     # initialize the venv
renv::activate() # activate the virtual environment
package.install("tidyverse") # install the most important package
renv::snapshot() # freeze packages in the lock file
```

For Python, 2-4 are nice to have, not mandatory:

1. conda - package manager |> environments
2. pipx - global packages `brew install pipx` |> tool for global tools
3. poetry - dependency manager |> packaging own libraries | lock dependencies
4. VSCode -> python, pylance extensions, jupyter
  - Alternatively, JupyterLab

```bash
# check that the python setup works
conda create -n my-venv-name python=3.10.8  # init/create
conda activate my-venv-name  # activate
conda install pandas         # install packages
conda install ipykernel      # for vscode and eventually jupyter notebooks/lab
```

For data:
1. Make an account in [Kaggle](https://www.kaggle.com/)
2. Install the Kaggle CLI


```{r}
mtcars |> head()
```